---
title: "DATA 606 - Lab0: Introduction to R and RStudio"
author: "Walt Wells, Fall 2016"
output:
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

```{r}

### Load Data
source("http://www.openintro.org/stat/data/arbuthnot.R")
source("http://www.openintro.org/stat/data/present.R")
```

## Explore Birth Records: 1940:2002

#### What years are included in this data set? What are the dimensions of the data frame and what are the variable or column names?

```{r, comment=NA}
c(head(present$year, n=1), tail(present$year, n=1))
dim(present)
names(present)
```

#### How do these counts compare to Arbuthnot's? Are they on a similar scale?

```{r, comment=NA}
## Let's calculate basic summary stats on the counts by year for each dataset
AB <- c(mean(arbuthnot$boys + arbuthnot$girls), median(arbuthnot$boys + arbuthnot$girls))
PT <- c(mean(present$boys + present$girls), median(present$boys + present$girls))
df <- data.frame(AB, PT, row.names = c("Mean", "Median"))
df$LogScale <- log(df$PT, df$AB)
df

```

We can see from the above table that the counts between the two data sets are quite different in scale.   AB represents the historical data, PT represents the modern data of which there is considerably more.

#### Make a plot that displays the boy-to-girl ratio for every year in the data set. What do you see? Does Arbuthnot's observation about boys being born in greater proportion than girls hold up in the U.S.? Include the plot in your response.

```{r, comment=NA}
plot(present$year, present$boys/present$girls, xlab="Year", ylab="Ratio of Boys to Girls", main="Ratio of Boy to Girl Births, 1940-2002")
sum(present$girls > present$boys)
```

We can see that during this period, the proportion of boys born is greater than the proportion of girls born.   Dr. Arbuthnot's observation is still holding true, but we can see that the ratio is trending in girls' favor.  

We briefly confirm our observation by checking briefly if there were any rows where girls outnumbered the boys.  

#### In what year did we see the most total number of births in the U.S.? 

```{r, comment=NA}
present$total <- present$boys + present$girls
maxyr <- present$year[(which(present$total==max(present$total)))]
max <- max(present$total)                     
```

We can see that the year with the maximum total births in the US was `r maxyr` with `r max` total births.

